---
title: "The Care Board Master Methodology and Pipeline"
subtitle: "Kansas Population Center, Institute for Policy & Social Research, University of Kansas"
format:
  pdf:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
params:
  execute_pipeline: false
  run_ipums_download: false
  run_database_upload: false
---

# Purpose

This master document assembles the available Care Board methodology and code. The default `execute_pipeline: false` includes the available analyses for documentation without recalculating their outputs. Missing sections are labeled explicitly. All 14 current pipeline documents and their required helpers are present. See [Reproducibility](docs/REPRODUCIBILITY.md) for external source inputs and verification limits, and [Software environment](docs/ENVIRONMENT.md) for captured dependency versions and setup.

After restoring and verifying those inputs, setting `execute_pipeline: true` evaluates the analytic QMD files in dependency order and writes updated outputs to `app_data/` and the replication folders under `data/`. The master stops before executing the analyses when a selected pipeline script is missing. Run commands from the repository root; `_quarto.yml` keeps computation paths relative to that root and writes rendered documents to `generated/`.

This version adds the Household Bargaining Power metrics and replaces the previous care ratio panel with the Care Resource Equity Score (CaRES). The Distribution of Care workflow now produces a combined Care Gini, a combined Care Location Quotient (LQ), and a bounded 0-1 CaRES score in `metrics_state_care_gini.csv`; the database upload uses the Gini and CaRES fields while retaining LQ in the CSV export for analysis.

Two steps are intentionally opt-in. `IPUMS_API.qmd` submits external IPUMS extract jobs and should be run only when the raw extracts need to be refreshed. `_database_upload.qmd` deletes and reloads warehouse tables and should be run only by a credentialed maintainer immediately before publication. Set `run_ipums_download: true` or `run_database_upload: true` only when those actions are intended.

# Required Inputs

For a full refresh from raw IPUMS extracts and supporting source files, place these files in the release folder before rendering this master document.

| Path | Role |
|---|---|
| `data/IPUMS Pulls/cps_00453.xml` and matching `.dat.gz` | Raw monthly CPS extract referenced by the recovered `data_processing.qmd`. |
| `data/IPUMS Pulls/cps_00452.xml` and matching `.dat.gz` | Raw CPS ASEC extract referenced by the recovered `data_processing.qmd`, including the Maternal Bargaining Power workflow. |
| `data/IPUMS Pulls/atus_00035.xml` and matching `.dat.gz` | Raw ATUS extract used by `data_processing.qmd`, including the Care Privilege workflow. |
| `data/IPUMS Pulls/atus_00029.xml` and matching `.dat.gz` | ATUS relationship extract used by `market_datum.qmd`. |
| `data/IPUMS Pulls/atus_00036.xml` and matching `.dat.gz` | Eldercare recipient roster attached during preprocessing. Its prior-three-month reference period differs from diary-day care time; see `docs/ATUS_ELDERCARE.md`. |
| `data/CSV/FormalOccs_Crossover.csv` | Paid-care occupation crosswalk. |
| `data/CSV/ATUSActivityCrossover.csv` | Unpaid-care activity crosswalk. |
| `data/CSV/GdpByState.csv` | BEA state GDP input for broad impacts valuation. |
| `data/CSV/Informal_Formal_Crosswalk.csv` | Replacement-wage crosswalk between unpaid activities and formal occupations. |
| `data/CSV/workforce_area_characteristics.csv` | LODES workplace-area characteristics input used to measure care employment for Care Gini, LQ, and CaRES. |
| `data/CSV/nhgis_populationbytract.csv` or `data/CSV/nighis_populationbytract.csv` | IPUMS NHGIS ACS 5-year tract population input used as the population denominator for Care Gini, LQ, and CaRES. The workflow accepts both the intended spelling and the currently uploaded spelling. |
| `data/CSV/resident_area_characteristics.csv` | LODES resident-area characteristics file retained with the GINI paper documentation; it is not required by the current state-level Care Board CaRES calculation. |
| `data/CSV/Tractlevel_Employment.csv` | Legacy tract-level care employment file retained as a fallback input for older care-ratio/Gini workflows. |
| `local_only/research/care_privilege_atus_privilege_focus.docx` | Local research reference for Care Privilege. Publication status has not been established, so the Word document is excluded from Git. The public implementation is in `analysis/bargainin_power.qmd`; this Word file is not read by the calculations. |
| `data/CSV/ASECSampleIDs.csv` | Optional input for `IPUMS_API.qmd` when requesting a fresh CPS ASEC extract. |
| `data/CSV/CPSSampleIDs.csv` | Optional input for `IPUMS_API.qmd` when requesting a fresh monthly CPS extract. |

If you are not reprocessing raw IPUMS extracts, provide the cleaned files generated by `data_processing.qmd`: `data/CSV/CPSdata.csv`, `data/CSV/ASECdata.csv`, and `data/CSV/ATUSdata.csv`.

The `app_data/` folder also contains static app metadata used during provider and database upload steps: `provider_group.xlsx`, `provider_category.xlsx`, `metric_tables.xlsx`, `metric_labels.csv`, and `source.csv`. Keep those files in `app_data/` unless the website metadata structure changes.

# Updated Metric Files

The app-facing metric files below are the main release outputs affected by the current update.

| File | Measures | Notes |
|---|---|---|
| `app_data/metrics_state_care_gini.csv` | `gini`, `lq`, `cares` | Replaces the earlier care-gini-only output. `gini` is the combined Care Gini, `lq` is the combined Care LQ retained in the CSV export, and `cares` is the 0-1 scaled CaRES score used in place of the old care ratio panel during upload. |
| `app_data/metrics_maternal_power.csv` | `share_of_income` | Adds Maternal Bargaining Power for married mother-father households with children, reported by year and geography. |
| `app_data/metrics_priviledge.csv` | `care_privileged_population`, `care_privileged_population_proportion` | Adds Care Privilege counts and proportions by year and geography. The file stem keeps the requested spelling used by the app workflow. |
| `app_data/metrics_sandwich_generation.csv` | Sandwich generation counts and proportions | Continues to support the Distribution of Care page alongside the new bargaining-power section. |

`analysis/legacy/care_ratio.qmd` and any prior `metrics_care_ratio` outputs are legacy artifacts in this release. The active database upload maps the scaled `cares` field from `metrics_state_care_gini.csv` into the existing care-ratio metric slot so the website infrastructure can display the replacement panel without a schema change.

# Run Order

```{r master-setup, include=FALSE}
options(knitr.duplicate.label = "allow")
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

strip_yaml <- function(lines) {
  if (length(lines) >= 2 && trimws(lines[1]) == "---") {
    yaml_end <- which(trimws(lines[-1]) == "---")[1] + 1
    if (!is.na(yaml_end)) {
      return(lines[-seq_len(yaml_end)])
    }
  }
  lines
}

demote_headings <- function(lines) {
  in_chunk <- FALSE
  vapply(lines, function(line) {
    if (grepl("^```", line)) {
      in_chunk <<- !in_chunk
      return(line)
    }
    if (!in_chunk && grepl("^#{1,5}\\s+", line)) {
      return(paste0("#", line))
    }
    line
  }, character(1))
}

force_eval_false <- function(lines) {
  out <- character()
  for (line in lines) {
    out <- c(out, line)
    if (grepl("^```\\{", line)) {
      out <- c(out, "#| eval: false")
    }
  }
  out
}

read_child_body <- function(file, execute = TRUE) {
  lines <- readLines(file, warn = FALSE)
  lines <- strip_yaml(lines)
  lines <- demote_headings(lines)
  if (!isTRUE(execute)) {
    lines <- force_eval_false(lines)
  }
  paste(lines, collapse = "\n")
}

render_child_qmd <- function(file, title, execute = TRUE) {
  cat("\n\n\\newpage\n\n")
  cat("# ", title, "\n\n", sep = "")
  if (!file.exists(file)) {
    cat("*Missing section: `", file, "` is not included in this repository snapshot. No methodology or output regeneration is claimed for this section.*\n\n", sep = "")
    return(invisible(NULL))
  }
  cat(
    if (execute) {
      "*Execution status: code chunks are evaluated in this master render.*\n\n"
    } else {
      "*Execution status: code chunks are included for documentation but not evaluated in this master render.*\n\n"
    }
  )

  child_env <- new.env(parent = globalenv())
  child_text <- read_child_body(file, execute = execute)
  child_output <- knitr::knit_child(text = child_text, envir = child_env, quiet = TRUE)
  cat(child_output, sep = "\n")
}

pipeline_files <- data.frame(
  file = paste0("analysis/", c(
    "IPUMS_API.qmd", "data_processing.qmd", "market.qmd", "market_datum.qmd",
    "activity_formal.qmd", "activity_informal.qmd", "care_provider.qmd",
    "provider.qmd", "broad_impacts.qmd", "gini.qmd", "sandwich_generation.qmd",
    "bargainin_power.qmd", "parental_labor_monitor.qmd", "_database_upload.qmd"
  )),
  title = c(
    "Raw IPUMS Download", "Data Processing and Harmonization", "Market Age Baseline",
    "Market Need and Provision Estimates", "Paid Care Activity Statistics",
    "Unpaid Care Activity Statistics", "Care Provider Flow Statistics",
    "Provider Demographic Statistics", "Broad Impacts Statistics",
    "Care Resource Equity Statistics", "Sandwich Generation Statistics",
    "Household Bargaining Power Statistics", "Parental Labor Monitor",
    "Database Upload Documentation"
  ),
  execute = c(
    isTRUE(params$execute_pipeline) && isTRUE(params$run_ipums_download),
    rep(isTRUE(params$execute_pipeline), 12),
    isTRUE(params$execute_pipeline) && isTRUE(params$run_database_upload)
  ),
  stringsAsFactors = FALSE
)

missing_selected <- pipeline_files$file[pipeline_files$execute & !file.exists(pipeline_files$file)]
if (length(missing_selected)) {
  stop("Selected pipeline scripts are missing: ", paste(missing_selected, collapse = ", "),
       ". See docs/REPRODUCIBILITY.md. No child analysis has run.", call. = FALSE)
}
if (any(pipeline_files$execute)) {
  source("R/load_defaults.R")
}
```

::: callout-note
#### Code Explanation

This chunk defines the document-assembly helpers and ordered pipeline. It checks selected scripts before any child analysis runs. Only an execution-enabled build sources `R/load_defaults.R`; a documentation-only build uses base R and `knitr` and does not install the analytic packages.
:::


```{r run-order-table, echo=FALSE}
run_table <- data.frame(
  File = pipeline_files$file,
  Section = pipeline_files$title,
  Availability = ifelse(file.exists(pipeline_files$file), "Present", "Missing"),
  Execution = ifelse(pipeline_files$execute, "Runs", "Documentation only")
)
knitr::kable(run_table)
```

::: callout-note
#### Code Explanation

This table shows which scripts are available and whether their code is selected for execution. A documentation-only render labels missing sections rather than claiming that their calculations were reproduced.
:::


# Methodology and Code

```{r run-children, results='asis'}
for (i in seq_len(nrow(pipeline_files))) {
  render_child_qmd(
    file = pipeline_files$file[[i]],
    title = pipeline_files$title[[i]],
    execute = pipeline_files$execute[[i]]
  )
}
```

::: callout-note
#### Code Explanation

The iteration step loops across years, states, datasets, categories, or demographic groups and binds the repeated results into a single analysis table. The main objects created or updated here are `file`, `title`, `execute`.
:::


# Generated Output Inventory

The table below lists app-facing files currently present after the render. In a full execution, these should reflect the newly generated outputs from the pipeline.

```{r output-inventory, echo=FALSE}
if (dir.exists("app_data")) {
  app_outputs <- list.files("app_data", pattern = "\\.(csv|xlsx)$", full.names = FALSE)
  output_tbl <- data.frame(
    file = sort(app_outputs),
    stringsAsFactors = FALSE
  )
  knitr::kable(output_tbl)
} else {
  cat("`app_data/` was not found.")
}
```

::: callout-note
#### Code Explanation

It imports source data into memory using the survey, spreadsheet, database, directory, or API reader shown in the code. The display code builds a diagnostic plot, publication figure, static table, or interactive HTML table from the objects created earlier. The main objects created or updated here are `app_outputs`, `output_tbl`, `file`, `stringsAsFactors`.
:::
