---
title: "02_data_download"
format: html
editor: visual
---

## Overview

-   This document provides instructions for downloading the **CPS** and **ASEC** data. Once these files are downloaded, move them into the **data** folder

-   Important Notes:

    -   **ATUS Data**: Download from github

    -   **Sample Definitions**: Before running the following steps, review the following files:

        -   `01_ASECSampleIDs.csv`

        -   `01_CPSSampleIDs.csv`

        You may need to add new rows to define samples for recent time periods.

    -   Use the following link to update the CPS sample IDs: <https://cps.ipums.org/cps-action/samples/sample_ids>.

## Download Setup

```{r Samples}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(
  ipumsr, 
  tidyverse
)

# List of samples to download
samples_asec <- read.csv("../01_preliminary-code-and-data/01_ASECSampleIDs.csv")$ASEC_Sample_IDs

samples_cps <- read.csv("../01_preliminary-code-and-data/01_CPSSampleIDs.csv")$CPS_Sample_IDs 
```

```{r Variables}
# Variables to extract
var_common <- c(
  "PERNUM",
  "REGION",
  "STATEFIP",
  "SEX",
  "RACE",
  "HISPAN",
  "AGE",
  "EDUC",
  "MARST",
  "MOMLOC",
  "POPLOC",
  "FAMSIZE",
  "NCHILD",
  "YNGCH",
  "IND1990",
  "OCC2010",
  "EMPSTAT",
  "WKSTAT",
  "AHRSWORKT",
  "ABSENT",
  "WHYABSNT"
)

var_asec <- c(
  var_common,
  "EARNWT",
  "UHRSWORKT",
  "INCWAGE",
  "POVERTY"
)

var_cps <- c(
  var_common,
  "LABFORCE",
  "SPLOC",
  "COMPWT",
  "TELWRKPAY",
  "CLASSWKR",
  "DIFFCARE",
  "NILFACT"  
)
```

## Data Download

```{r Download}
# update the following if you only want to download a specific data set list
data_sets <- c('asec', 'cps')

for(sel_data_set in data_sets){

  # Create data extract
  cps_ext_def <- define_extract_cps(
    description = "Care Variable Extract",
    samples = get(paste0('samples_', sel_data_set)),
    variables = get(paste0('var_', sel_data_set))
  )
  Sys.sleep(60)

  # Submit data extract request
  cps_ext_submitted <- submit_extract(cps_ext_def)
  cps_ext_complete <- wait_for_extract(cps_ext_submitted)

  # Download extract
  filepath <- download_extract(cps_ext_submitted, overwrite = TRUE)
  Sys.sleep(60)

  print(paste(sel_data_set, 'data has been downloaded'))
}

print("API work has finished")
```
