---
title: "Care Ratio Methodology"
format:
  html:
    toc: true
    toc-depth: 3
    code-overflow: wrap
    df-print: paged
  pdf:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
  echo: true
editor: visual
date: 2025-08-07
---

# Methods Overview

This file builds the care-ratio metric used by The Care Board to compare potential demand for care with the local capacity of care-related employment. Conceptually, the measure asks whether the places where people are likely to need care align with the places where care-sector jobs are located.

The workflow combines tract-level employment data, age and disability information, and the market-need estimates prepared elsewhere in the project. Employment in education, health care, and related care industries is used as a proxy for formal care capacity. Population characteristics are used to weight expected care need. The final metric is written at state and national levels so it can be displayed with other broad impacts measures on the website.

**Outputs.** This workflow writes `metrics_care_ratio` to `app_data/` for the website and to `data/CSV/`, `data/Dta/`, and `data/Excel/` for replication.

**Key sources.** The geographic employment approach draws on tract-level employment and Census/ACS-style small-area tabulations. The file uses `tidycensus` for Census API access and draws on care-need assumptions produced from ATUS and ASEC in the market files. The spatial employment foundation is related to the Census [LEHD/LODES](https://lehd.ces.census.gov/data/) approach to workplace-area employment data.

## Introduction and Literature Context

Care access is spatial. A state can have many care jobs in the aggregate while particular counties or neighborhoods still face shortages relative to likely need. The care-ratio metric compares expected care demand with care-related employment capacity, using age, disability, population, and employment measures to summarize whether care resources are aligned with community need.

The statistic is intentionally a planning indicator rather than a direct service-access measure. It does not observe every care provider or every household's unmet need. Instead, it provides a transparent signal about geographic imbalance that can be compared over time and across states.

## Methodological Decisions

| Decision | Implementation | Rationale |
|---|---|---|
| Capacity proxy | Use education, health care, and related care-industry employment | Captures local formal-care infrastructure available in employment data. |
| Need proxy | Combine age/disability ACS measures with market-derived care weights | Reflects life-course and functional need differences. |
| API access | Prompt for a Census API key at runtime | Keeps public code credential-free while allowing reproducible ACS pulls. |
| Output scale | Produce state and national metrics | Aligns with the Broad Indicators display structure. |

## Policy Implications

The care ratio can help agencies identify places where care infrastructure may not match demographic need. It is relevant for targeted workforce incentives, child care and long-term care investments, transportation and service-area planning, and regional equity analysis. Because the measure is spatial, it can highlight access problems that statewide averages hide.

```{r CR step_1_01, message=FALSE, warning=FALSE}
library(dplyr)
library(DescTools)
library(lctools)
library(tidycensus)
library(tidyr)
library(doParallel)
library(foreach)
library(purrr)

source("R/load_defaults.R")

DF <- read.csv('./data/CSV/Tractlevel_Employment.csv')
DF$GEOID <- sprintf("%011s", as.character(DF$GEOID))
DF <- DF %>%  # pipe
# Create or modify columns
mutate(
    Educational_Services = ifelse(is.na(Educational_Services), 0, Educational_Services),
    Health_Care = ifelse(is.na(Health_Care), 0, Health_Care),
    Others = ifelse(is.na(Others), 0, Others)
  )
DF <- DF |>  # pipe
# Create or modify columns
mutate(
    Care_Industry = Educational_Services + Health_Care + Others
  )

DF$STATEFP <- substr(DF$GEOID, 1, 2)
DF$COUNTYFP <- substr(DF$GEOID, 3, 5)
DF$STATEFP <- sprintf("%02d", as.numeric(DF$STATEFP))


States <- unique(DF$STATEFP)

register_census_api_key()


years <- 2014:2021

employment_var <- "B23025_004"  # Civilian employed population 16+
```

::: callout-note
#### Code Explanation

This chunk loads `dplyr`, `DescTools`, `lctools`, `tidycensus`, `tidyr`, `doParallel`, `foreach`, `purrr` package(s) and sources `R/load_defaults.R` so the later code has the required functions, defaults, and shared helpers. This chunk registers an API key through the shared terminal-prompt helper, which keeps credentials out of the public QMD while still allowing the data pull to run interactively. It reads `./data/CSV/Tractlevel_Employment.csv` into memory and prepares those inputs for the transformations that follow. Technically, it uses `mutate()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics.
:::


## Loop over years and bind rows

```{r CR loop_over_years_and_bind_rows_02, message=FALSE, warning=FALSE}
# Loop over years and bind rows
employment_all_years <- map_dfr(years, function(y) {
  get_acs(
    geography = "county",
    variables = employment_var,
    year = y,
    survey = "acs5",
    output = "wide"
  ) %>%  # pipe
# Create or modify columns
mutate(year = y)
})
```

::: callout-note
#### Code Explanation

It imports source data into memory using the survey, spreadsheet, database, directory, or API reader shown in the code. Technically, it uses `mutate()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. It defines reusable function logic so repeated calculations can be applied consistently across groups, years, or geographies. The iteration step loops across years, states, datasets, categories, or demographic groups and binds the repeated results into a single analysis table.
:::


## Optional: Keep only estimate (not MOE)

```{r CR optional_keep_only_estimate_not_moe_03, message=FALSE, warning=FALSE}
# Optional: Keep only estimate (not MOE)
employment_estimates <- employment_all_years %>%  # pipe
# Select/reorder columns
select(GEOID, NAME, year, B23025_004E) %>%  # pipe
  rename(employed = B23025_004E)


employment_estimates$STATEFP <- substr(employment_estimates$GEOID, 1, 2)
employment_estimates$COUNTYFP <- substr(employment_estimates$GEOID, 3, 5)
```

::: callout-note
#### Code Explanation

Technically, it uses `select()`, `rename()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `employment_estimates`.
:::


## Define all variable codes

```{r CR define_all_variable_codes_04, message=FALSE, warning=FALSE}
# Define all variable codes
vars <- c(
  total_population = "B01003_001",
  male_under5 = "B01001_003",
  female_under5 = "B01001_027",
  male_5_9 = "B01001_004",
  female_5_9 = "B01001_028",
  male_10_14 = "B01001_005",
  female_10_14 = "B01001_029",
  male_15_17 = "B01001_006",
  female_15_17 = "B01001_030",
  male_75_79 = "B01001_023",
  male_80_84 = "B01001_024",
  male_85_up = "B01001_025",
  female_75_79 = "B01001_047",
  female_80_84 = "B01001_048",
  female_85_up = "B01001_049",
  total_disab = "C18108_001",
  disab_total = "C18108_002"
)
```

::: callout-note
#### Code Explanation

The main objects created or updated here are `vars`, `total_population`, `male_under5`, `female_under5`, `male_5_9`.
:::


## Function to retrieve and process data

```{r CR function_to_retrieve_and_process_data_05, message=FALSE, warning=FALSE}
# Function to retrieve and process data
get_county_demo <- function(y) {
  df <- tryCatch({
    get_acs(
      geography = "county",
      variables = vars,
      year = y,
      survey = "acs5",
      output = "tidy"
    ) %>%  # pipe
# Create or modify columns
mutate(year = y)
  }, error = function(e) {
    message(paste("Skipping year", y, "due to error:", e$message))
    return(NULL)
  })
  return(df)
}
```

::: callout-note
#### Code Explanation

It imports source data into memory using the survey, spreadsheet, database, directory, or API reader shown in the code. Technically, it uses `mutate()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. It defines `get_county_demo()`, which encapsulates repeated logic used by later chunks. The main objects created or updated here are `get_county_demo`, `df`, `geography`, `variables`, `year`.
:::


## Get data for all years

```{r CR get_data_for_all_years_06, message=FALSE, warning=FALSE}
# Get data for all years
county_demog <- map_dfr(years, get_county_demo)

county_demog$STATEFP <- substr(county_demog$GEOID, 1, 2)
county_demog$COUNTYFP <- substr(county_demog$GEOID, 3, 5)
```

::: callout-note
#### Code Explanation

The iteration step loops across years, states, datasets, categories, or demographic groups and binds the repeated results into a single analysis table. The main objects created or updated here are `county_demog`.
:::


## To understand the weights of different populations we need to use the market_Datum data

```{r CR to_understand_the_weights_of_different_populations_we_need_to_use_the_market_datum_data_07, message=FALSE, warning=FALSE}
#To understand the weights of different populations we need to use the market_Datum data
market_datum <- fread("./app_data/market_datum.csv") |>  # pipe 
# Group rows for downstream summaries
group_by(age) |>  # pipe
# Compute summary statistics
summarise(
    need = sum(need_interval)
  ) %>%  # pipe
  #Weight refers to for each age group, on a scale of 0-1 how much care is needed with 1 representing the highest need group.
# Create or modify columns
mutate(
    weight = need/max(need)
  )
```

::: callout-note
#### Code Explanation

It reads `./app_data/market_datum.csv` into memory and prepares those inputs for the transformations that follow. Technically, it uses `mutate()`, `group_by()`, `summarise()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The statistical work in this chunk calculates weighted counts, rates, rolling values, medians, changes over time, or inequality measures used in the published indicators. The main objects created or updated here are `market_datum`, `need`, `weight`.
:::


## We normalize this weight creating a mean of 1 and a stnadard deviation of 1

```{r CR we_normalize_this_weight_creating_a_mean_of_1_and_a_stnadard_deviation_of_1_08, message=FALSE, warning=FALSE}
#We normalize this weight creating a mean of 1 and a stnadard deviation of 1
market_datum$weight <- (market_datum$weight - mean(market_datum$weight, na.rm = TRUE)) / sd(market_datum$weight, na.rm = TRUE) + 1

data <- read.csv("./data/CSV/CPSdata.csv") |>  # pipe
# Filter rows by conditions
filter(YEAR >= 2014) |>  # pipe
# Filter rows by conditions
filter(nilf_activity == "Homemaker") |>  # pipe
# Select/reorder columns
select(YEAR, month, nilf_activity, WTFINL, STATEFIP)

data$STATEFIP <- sprintf("%02d", as.numeric(data$STATEFIP))
```

::: callout-note
#### Code Explanation

It reads `./data/CSV/CPSdata.csv` into memory and prepares those inputs for the transformations that follow. Technically, it uses `filter()`, `select()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The statistical work in this chunk calculates weighted counts, rates, rolling values, medians, changes over time, or inequality measures used in the published indicators. The main objects created or updated here are `data`.
:::


## DENOMINATOR

```{r CR denominator_09, message=FALSE, warning=FALSE}
##### DENOMINATOR

Denominators = {}
Years = {}
State = {}

for (st in States){
  temp_demo_st <- county_demog |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
for (yr in years) {
  temp_demo <- temp_demo_st |>  # pipe
# Filter rows by conditions
filter(year == yr) 
  
  under5 <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_under5" | variable == "female_under5")
  under5 <- sum(under5$estimate)
  
  under5_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age < 5)
  under5_W <- mean(under5_W$weight)
  
  five_nine <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_5_9" | variable == "female_5_9")
  five_nine <- sum(five_nine$estimate)
    
  five_nine_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 4 & age < 10)
  five_nine_W <- mean(five_nine_W$weight)
  
  ten_fourteen <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_10_14" | variable == "female_10_14")
  ten_fourteen <- sum(ten_fourteen$estimate)
  
  ten_fourteen_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 9 & age < 15)
  ten_fourteen_W <- mean(ten_fourteen_W$weight)
  
  fifteen_seventeen <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_15_17" | variable == "female_15_17")
  fifteen_seventeen <- sum(fifteen_seventeen$estimate)
  
  fifteen_seventeen_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 14 & age < 18)
  fifteen_seventeen_W <- mean(fifteen_seventeen_W$weight)
  
  seventies <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_75_79" | variable == "female_75_79")
  seventies <- sum(seventies$estimate)
  
  seventies_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 74 & age < 80)
  seventies_W <- mean(seventies_W$weight)
  
  eighties <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_80_84" | variable == "female_80_84")
  eighties <- sum(eighties$estimate)
  
  eighties_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 79 & age < 85)
  eighties_W <- mean(eighties_W$weight)
  
  over85 <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_85_up" | variable == "female_85_up")
  over85 <- sum(over85$estimate)
  
  over85_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 84)
  over85_W <- mean(over85_W$weight)
  
  disabled <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "total_disab")
  disabled <- sum(disabled$estimate)
  
  disableW <- max(market_datum$weight)
  
  Denom = under5*under5_W +
    five_nine*five_nine_W +
    ten_fourteen*ten_fourteen_W +
    fifteen_seventeen*fifteen_seventeen_W +
    seventies*seventies_W +
    eighties*eighties_W +
    over85*over85_W
  
  Denominators = append(Denominators, Denom)
  Years = append(Years, yr)
  State = append(State, st)
}
}
```

::: callout-note
#### Code Explanation

Technically, it uses `filter()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The iteration step loops across years, states, datasets, categories, or demographic groups and binds the repeated results into a single analysis table. The statistical work in this chunk calculates weighted counts, rates, rolling values, medians, changes over time, or inequality measures used in the published indicators. The main objects created or updated here are `Denominators`, `Years`, `State`, `temp_demo_st`, `temp_demo`.
:::


## NUMERATORS

```{r CR numerators_10, message=FALSE, warning=FALSE}
### NUMERATORS

Numerators = {}
Years = {}
State = {}
W = c(1.5, 0.5, 1)

for (st in States) {
  temp_st <- DF |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
  temp_demo_st <- county_demog |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
  temp_employ_st <- employment_estimates |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
  cps_temp_st <- data |>  # pipe
# Filter rows by conditions
filter(STATEFIP == st)
  for (yr in years) {
    temp <- temp_st |>  # pipe
# Filter rows by conditions
filter(year == yr)
    
    temp_employ <- temp_employ_st |>  # pipe
# Filter rows by conditions
filter(year == yr)
    
    temp_demo <- temp_demo_st |>  # pipe
# Filter rows by conditions
filter(year == yr)
    
    cps_temp <- cps_temp_st %>%  # pipe
# Filter rows by conditions
filter(YEAR == yr)
    
    population <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "total_population")
    population <- sum(population$estimate)
    
    careworkers <- sum(temp$Care_Industry, na.rm = TRUE)
    
    workingnoncare <- sum(temp_employ$employed)-careworkers
      
    Homemakers <- sum(cps_temp$WTFINL/length(unique(cps_temp$month)))
    
    Years = append(Years, yr)
    print(yr)
    State = append(State, st)
    print(st)
    Numer <- careworkers*W[1] + workingnoncare*W[2] + Homemakers*W[3]
    print(Numer)
    Numerators<- append(Numerators, Numer)
  }
}

care_ratio <- data.frame(
  date = as.Date(paste0(Years, "-01-01")), 
  care_ratio = Numerators / Denominators,
  geo_level = State
)

care_ratio$geo_level <- paste0("state", care_ratio$geo_level)
```

::: callout-note
#### Code Explanation

Technically, it uses `filter()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The iteration step loops across years, states, datasets, categories, or demographic groups and binds the repeated results into a single analysis table. The statistical work in this chunk calculates weighted counts, rates, rolling values, medians, changes over time, or inequality measures used in the published indicators. The main objects created or updated here are `Numerators`, `Years`, `State`, `W`, `temp_st`.
:::


## National

```{r CR national_11, message=FALSE, warning=FALSE}
# National

cr <- care_ratio |>  # pipe
# Group rows for downstream summaries
group_by(date) |>  # pipe
# Compute summary statistics
summarise(care_ratio = mean(care_ratio, na.rm = TRUE)) |>  # pipe
# Create or modify columns
mutate(geo_level = "national")

care_ratio <- rbind(care_ratio,cr)

write_careboard_outputs(
  care_ratio,
  "metrics_care_ratio",
  app_path = "./app_data/metrics_care_ratio.csv"
)
```

::: callout-note
#### Code Explanation

Technically, it uses `mutate()`, `group_by()`, `summarise()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The statistical work in this chunk calculates weighted counts, rates, rolling values, medians, changes over time, or inequality measures used in the published indicators. It writes the resulting table through the shared export helper, producing app-facing files and/or replication copies in CSV, Stata, and Excel formats. The main objects created or updated here are `cr`, `care_ratio`, `app_path`.
:::
