---
title: "care_ratio — annotated"
format:
  html:
    toc: true
    toc-depth: 3
    code-overflow: wrap
    df-print: paged
execute:
  warning: false
  message: false
  echo: true
editor: visual
date: 2025-08-07
---

# Methods Overview

```{r step_1_01, message=FALSE, warning=FALSE}
library(dplyr)
library(DescTools)
library(lctools)
library(tidycensus)
library(tidyr)
library(doParallel)
library(foreach)
library(purrr)

source("load_defaults.R")

DF <- read.csv('./data/CSV/Tractlevel_Employment.csv')
DF$GEOID <- sprintf("%011s", as.character(DF$GEOID))
DF <- DF %>%  # pipe
# Create or modify columns
mutate(
    Educational_Services = ifelse(is.na(Educational_Services), 0, Educational_Services),
    Health_Care = ifelse(is.na(Health_Care), 0, Health_Care),
    Others = ifelse(is.na(Others), 0, Others)
  )
DF <- DF |>  # pipe
# Create or modify columns
mutate(
    Care_Industry = Educational_Services + Health_Care + Others
  )

DF$STATEFP <- substr(DF$GEOID, 1, 2)
DF$COUNTYFP <- substr(DF$GEOID, 3, 5)
DF$STATEFP <- sprintf("%02d", as.numeric(DF$STATEFP))


States <- unique(DF$STATEFP)

census_api_key("bc4a1c0184ce6c385720174b31b312e5a703262d", install = TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")


years <- 2014:2021

employment_var <- "B23025_004"  # Civilian employed population 16+
```

## Loop over years and bind rows

```{r loop_over_years_and_bind_rows_02, message=FALSE, warning=FALSE}
# Loop over years and bind rows
employment_all_years <- map_dfr(years, function(y) {
  get_acs(
    geography = "county",
    variables = employment_var,
    year = y,
    survey = "acs5",
    output = "wide"
  ) %>%  # pipe
# Create or modify columns
mutate(year = y)
})
```

## Optional: Keep only estimate (not MOE)

```{r optional_keep_only_estimate_not_moe_03, message=FALSE, warning=FALSE}
# Optional: Keep only estimate (not MOE)
employment_estimates <- employment_all_years %>%  # pipe
# Select/reorder columns
select(GEOID, NAME, year, B23025_004E) %>%  # pipe
  rename(employed = B23025_004E)


employment_estimates$STATEFP <- substr(employment_estimates$GEOID, 1, 2)
employment_estimates$COUNTYFP <- substr(employment_estimates$GEOID, 3, 5)
```

## Define all variable codes

```{r define_all_variable_codes_04, message=FALSE, warning=FALSE}
# Define all variable codes
vars <- c(
  total_population = "B01003_001",
  male_under5 = "B01001_003",
  female_under5 = "B01001_027",
  male_5_9 = "B01001_004",
  female_5_9 = "B01001_028",
  male_10_14 = "B01001_005",
  female_10_14 = "B01001_029",
  male_15_17 = "B01001_006",
  female_15_17 = "B01001_030",
  male_75_79 = "B01001_023",
  male_80_84 = "B01001_024",
  male_85_up = "B01001_025",
  female_75_79 = "B01001_047",
  female_80_84 = "B01001_048",
  female_85_up = "B01001_049",
  total_disab = "C18108_001",
  disab_total = "C18108_002"
)
```

## Function to retrieve and process data

```{r function_to_retrieve_and_process_data_05, message=FALSE, warning=FALSE}
# Function to retrieve and process data
get_county_demo <- function(y) {
  df <- tryCatch({
    get_acs(
      geography = "county",
      variables = vars,
      year = y,
      survey = "acs5",
      output = "tidy"
    ) %>%  # pipe
# Create or modify columns
mutate(year = y)
  }, error = function(e) {
    message(paste("Skipping year", y, "due to error:", e$message))
    return(NULL)
  })
  return(df)
}
```

## Get data for all years

```{r get_data_for_all_years_06, message=FALSE, warning=FALSE}
# Get data for all years
county_demog <- map_dfr(years, get_county_demo)

county_demog$STATEFP <- substr(county_demog$GEOID, 1, 2)
county_demog$COUNTYFP <- substr(county_demog$GEOID, 3, 5)
```

## To understand the weights of different populations we need to use the market_Datum data

```{r to_understand_the_weights_of_different_populations_we_need_to_use_the_market_datum_data_07, message=FALSE, warning=FALSE}
#To understand the weights of different populations we need to use the market_Datum data
market_datum <- fread("./app_data/market_datum.csv") |>  # pipe 
# Group rows for downstream summaries
group_by(age) |>  # pipe
# Compute summary statistics
summarise(
    need = sum(need_interval)
  ) %>%  # pipe
  #Weight refers to for each age group, on a scale of 0-1 how much care is needed with 1 representing the highest need group.
# Create or modify columns
mutate(
    weight = need/max(need)
  )
```

## We normalize this weight creating a mean of 1 and a stnadard deviation of 1

```{r we_normalize_this_weight_creating_a_mean_of_1_and_a_stnadard_deviation_of_1_08, message=FALSE, warning=FALSE}
#We normalize this weight creating a mean of 1 and a stnadard deviation of 1
market_datum$weight <- (market_datum$weight - mean(market_datum$weight, na.rm = TRUE)) / sd(market_datum$weight, na.rm = TRUE) + 1

data <- read.csv("./data/CSV/CPSdata.csv") |>  # pipe
# Filter rows by conditions
filter(YEAR >= 2014) |>  # pipe
# Filter rows by conditions
filter(nilf_activity == "Homemaker") |>  # pipe
# Select/reorder columns
select(YEAR, month, nilf_activity, WTFINL, STATEFIP)

data$STATEFIP <- sprintf("%02d", as.numeric(data$STATEFIP))
```

## DENOMINATOR

```{r denominator_09, message=FALSE, warning=FALSE}
##### DENOMINATOR

Denominators = {}
Years = {}
State = {}

for (st in States){
  temp_demo_st <- county_demog |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
for (yr in years) {
  temp_demo <- temp_demo_st |>  # pipe
# Filter rows by conditions
filter(year == yr) 
  
  under5 <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_under5" | variable == "female_under5")
  under5 <- sum(under5$estimate)
  
  under5_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age < 5)
  under5_W <- mean(under5_W$weight)
  
  five_nine <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_5_9" | variable == "female_5_9")
  five_nine <- sum(five_nine$estimate)
    
  five_nine_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 4 & age < 10)
  five_nine_W <- mean(five_nine_W$weight)
  
  ten_fourteen <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_10_14" | variable == "female_10_14")
  ten_fourteen <- sum(ten_fourteen$estimate)
  
  ten_fourteen_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 9 & age < 15)
  ten_fourteen_W <- mean(ten_fourteen_W$weight)
  
  fifteen_seventeen <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_15_17" | variable == "female_15_17")
  fifteen_seventeen <- sum(fifteen_seventeen$estimate)
  
  fifteen_seventeen_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 14 & age < 18)
  fifteen_seventeen_W <- mean(fifteen_seventeen_W$weight)
  
  seventies <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_75_79" | variable == "female_75_79")
  seventies <- sum(seventies$estimate)
  
  seventies_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 74 & age < 80)
  seventies_W <- mean(seventies_W$weight)
  
  eighties <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_80_84" | variable == "female_80_84")
  eighties <- sum(eighties$estimate)
  
  eighties_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 79 & age < 85)
  eighties_W <- mean(eighties_W$weight)
  
  over85 <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "male_85_up" | variable == "female_85_up")
  over85 <- sum(over85$estimate)
  
  over85_W <- market_datum |>  # pipe
# Filter rows by conditions
filter(age > 84)
  over85_W <- mean(over85_W$weight)
  
  disabled <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "total_disab")
  disabled <- sum(disabled$estimate)
  
  disableW <- max(market_datum$weight)
  
  Denom = under5*under5_W +
    five_nine*five_nine_W +
    ten_fourteen*ten_fourteen_W +
    fifteen_seventeen*fifteen_seventeen_W +
    seventies*seventies_W +
    eighties*eighties_W +
    over85*over85_W
  
  Denominators = append(Denominators, Denom)
  Years = append(Years, yr)
  State = append(State, st)
}
}
```

## NUMERATORS

```{r numerators_10, message=FALSE, warning=FALSE}
### NUMERATORS

Numerators = {}
Years = {}
State = {}
W = c(1.5, 0.5, 1)

for (st in States) {
  temp_st <- DF |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
  temp_demo_st <- county_demog |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
  temp_employ_st <- employment_estimates |>  # pipe
# Filter rows by conditions
filter(STATEFP == st)
  cps_temp_st <- data |>  # pipe
# Filter rows by conditions
filter(STATEFIP == st)
  for (yr in years) {
    temp <- temp_st |>  # pipe
# Filter rows by conditions
filter(year == yr)
    
    temp_employ <- temp_employ_st |>  # pipe
# Filter rows by conditions
filter(year == yr)
    
    temp_demo <- temp_demo_st |>  # pipe
# Filter rows by conditions
filter(year == yr)
    
    cps_temp <- cps_temp_st %>%  # pipe
# Filter rows by conditions
filter(YEAR == yr)
    
    population <- temp_demo |>  # pipe
# Filter rows by conditions
filter(variable == "total_population")
    population <- sum(population$estimate)
    
    careworkers <- sum(temp$Care_Industry, na.rm = TRUE)
    
    workingnoncare <- sum(temp_employ$employed)-careworkers
      
    Homemakers <- sum(cps_temp$WTFINL/length(unique(cps_temp$month)))
    
    Years = append(Years, yr)
    print(yr)
    State = append(State, st)
    print(st)
    Numer <- careworkers*W[1] + workingnoncare*W[2] + Homemakers*W[3]
    print(Numer)
    Numerators<- append(Numerators, Numer)
  }
}

care_ratio <- data.frame(
  date = as.Date(paste0(Years, "-01-01")), 
  care_ratio = Numerators / Denominators,
  geo_level = State
)

care_ratio$geo_level <- paste0("state", care_ratio$geo_level)
```

## National

```{r national_11, message=FALSE, warning=FALSE}
# National

cr <- care_ratio |>  # pipe
# Group rows for downstream summaries
group_by(date) |>  # pipe
# Compute summary statistics
summarise(care_ratio = mean(care_ratio, na.rm = TRUE)) |>  # pipe
# Create or modify columns
mutate(geo_level = "national")

care_ratio <- rbind(care_ratio,cr)

write.csv(care_ratio, "./app_data/metrics_care_ratio.csv", row.names = FALSE)
```
