---
title: "market_datum"
format:
  html:
    toc: true
    code-overflow: wrap
    df-print: paged
  pdf:
    toc: true
execute:
  warning: false
  message: false
editor: visual
---

This workflow builds age-by-care-focus "market" estimates of caregiving need and caregiving provision by age and care focus. We utilize the data from ATUS to create the data informed estimates of the amount of care a person of a specific age group needs and the amount of care a person of a specific age group provides. It is important to note that there is certainly large variation in the amount of care needed and provided within an age group. We utilize data to create estimates of the average amount of time needed and provided, but based on individual characteristics, this amount will likely over estimate some people and under estimate others.

# Overview

This workflow builds age-by-care-focus market estimates of caregiving need and caregiving provision by age. It starts from the age distribution prepared in `market.qmd`, expands each age into the project's three care focus categories, and then uses ATUS activity records to estimate average minutes of need and provision.

These estimates are not intended to describe every person within an age group. They are population averages used to support website-level comparisons of care demand, care supply, and care-force geography. Some age and care-focus cells require project-defined overrides where survey data are sparse or where time-diary relationships do not fully capture expected care need.

**Outputs.** This workflow writes `market_datum` to `app_data/` for the website and to `data/CSV/`, `data/Dta/`, and `data/Excel/` for replication.

**Key sources.** The source microdata come from ATUS via IPUMS ATUS. Time-use surveys are widely used to measure unpaid care and domestic work; see the ILO discussion of [unpaid domestic and care work measurement](https://ilostat.ilo.org/topics/unpaid-work/measuring-unpaid-domestic-and-care-work/) and the BLS [ATUS calculation handbook](https://www.bls.gov/opub/hom/atus/calculation.htm).

## Introduction and Literature Context

The market-datum file translates demographic structure into care-market concepts: expected need, expected provision, and the balance between them by age and care focus. Time-use data are valuable here because they observe care-related activity in daily life rather than only counting paid service encounters. At the same time, time diaries are not clinical assessments of need. The estimates should therefore be read as population-level planning inputs, not as person-level diagnoses.

The approach follows a mixed empirical and judgment-based design. ATUS records provide the empirical foundation, while KU-defined overrides are used where survey data are sparse or where the diary relationship structure is known to understate expected care for certain ages. The document flags those decisions because they are substantive assumptions, not hidden data transformations.

## Methodological Decisions

| Decision | Implementation | Rationale |
|---|---|---|
| Care focus categories | Developmental, daily living, and health | Matches the Care Board conceptual model and downstream app categories. |
| ATUS window | Recent five-year window excluding 2020 | Improves stability and avoids pandemic-disrupted time-use data. |
| Relationship data | Join hierarchical ATUS relationship records | Helps identify care activity context and household relationships. |
| Overrides | Apply explicit age-by-focus assumptions where data are sparse | Prevents implausible zeros or unstable cells from driving public indicators. |
| Smoothing | Smooth age profiles while preserving bounds | Reduces noise across adjacent ages without erasing life-course patterns. |

## Policy Implications

Care-market estimates help translate demographic change into practical service questions: where need is likely to rise, which age groups provide care, and whether care demand and provision are balanced. These statistics can inform long-term care planning, child care capacity decisions, disability-service coordination, and workforce development strategies.

### Initialize Dataframe

The first chunk loads the project defaults and required packages and reads in the data from the previous section representing the age and population data which will act as the starting point of this analysis. We then expand the market data creating three rows for each age, one for each of the care focuses.

```{r market datum setup_01, message=FALSE, warning=FALSE}
source("R/load_defaults.R")

# Read age distribution prepared in the separate 'market' pipeline
age_modified <- read.csv("./app_data/market.csv")

# Create a base grid of all ages x care_focus categories
market_datum <- age_modified |>
  cross_join(data.frame(care_focus = c("developmental", "daily_living", "health")))
```

::: callout-note
#### Code Explanation

This chunk sources `R/load_defaults.R` so the later code has the required functions, defaults, and shared helpers. It reads `./app_data/market.csv` into memory and prepares those inputs for the transformations that follow. Technically, it uses `cross_join()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `age_modified`, `market_datum`.
:::


### Base ATUS setup

This chunk imports ATUS activity microdata which is used to estimate the need/provision intervals. This chunk isolates the most recent five years, excluding 2020, and constructs daily weights for the multi-year window.

```{r market datum data_import_02, message=FALSE, warning=FALSE}
# Read ATUS activity microdata (activity-level records)
atus <- read.csv("./data/CSV/ATUSdata.csv") %>%
  select(
    YEAR, CASEID, WT06, ACTLINE, HH_SIZE, AGE, marst, nchild,
    activity, act_care_focus, occ_care_focus,
    paid_work, child_care, elder_care, sleeping,
    DURATION, SCC_ALL_LN, SEC_ALL_LN
  ) %>%
  clean_names() %>%
  # Remove a known bad ID if required by data notes
  filter(caseid != 20220402221228)

# Keep the 5 most recent years, excluding 2020 (pandemic anomaly year)
years_include <- atus |>
  distinct(year) |>
  filter(year != 2020) |>
  arrange(desc(year)) |>
  slice_head(n = 5) |>
  pull()

# Restrict ATUS to selected years and prepare working variables
atus <- atus |>
  filter(year %in% years_include) |>
  rename(care_focus = act_care_focus) |>
  mutate(
    # care_job flags occupations with any care focus
    care_job  = ifelse(occ_care_focus == "none", 0, 1),
    # Convert ATUS annual weight to an average daily weight over 5 years
    weight    = wt06 / 365 / 5,
    # Work time attributable to paid care work (used in provision construction)
    work_time = duration * paid_work * care_job
  )
```

::: callout-note
#### Code Explanation

It reads `./data/CSV/ATUSdata.csv` into memory and prepares those inputs for the transformations that follow. Technically, it uses `filter()`, `mutate()`, `select()`, `rename()`, `clean_names()`, `arrange()`, `distinct()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `atus`, `years_include`, `care_job`, `weight`, `work_time`.
:::


### Incorporate Activity Relationship Data

This chunk merges ATUS household/person-relationship fields so that we can identify household composition and respondent role (i.e. living alone, reference person, etc). The ATUS file loaded in here is a heirarchical file that is used to understand who an activity is conducted with as opposed to just what the activity is. ATUS presents the data as both activity and heirarchical level data, we utilize both to create our estimates.

```{r market datum setup_03, message=FALSE, warning=FALSE}
# Read IPUMS ATUS Household/Person files (relationship fields)
library(ipumsr)

ddi_file <- read_ipums_ddi("./data/IPUMS Pulls/atus_00029.xml")
atus_hh   <- read_ipums_micro(ddi_file) |> clean_names()

# Join relationship info to activity rows via caseid + activity line
atus <- atus |>
  left_join(
    atus_hh |> select(caseid, actlinew, relatew),
    by = c("caseid" = "caseid", "actline" = "actlinew")
  )
```

::: callout-note
#### Code Explanation

This chunk loads `ipumsr` package(s) so the later code has the required functions, defaults, and shared helpers. It reads `./data/IPUMS Pulls/atus_00029.xml` into memory and prepares those inputs for the transformations that follow. Technically, it uses `select()`, `clean_names()`, `left_join()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `ddi_file`, `atus_hh`, `atus`, `by`.
:::


### Data Overrides

This chunk creates a function called prepare_overrides. This isn't important at this moment, but later on in the code for this section we will override certain deata results with our own assumptions. This function will be needed to accomplish this. We will describe the reaoning for this when we do it, but due to the non-availability of certain age groups in the data, we will need to make assumptions in certain points.

```{r market datum data_wrangling_04, message=FALSE, warning=FALSE}
prepare_overrides <- function(care_interval, age_ranges, col_name) {
  # Stack a list of named vectors: each element corresponds to an age_range;
  # each named entry corresponds to a care_focus with an interval value.
  care_override <- do.call(rbind, lapply(names(care_interval), function(age_range) {
    data <- care_interval[[age_range]]
    data.frame(
      age_range  = age_range,
      care_focus = names(data),
      interval   = as.numeric(data)
    )
  }))

  care_override <- care_override |>
    # Expand to single-year ages by joining the age-range map
    left_join(age_ranges, by = "age_range", relationship = "many-to-many") |>
    select(age, care_focus, interval) |>
    # Name the output column (e.g., "need_override" or "provision_override")
    rename({{ col_name }} := interval)

  care_override
}
```

::: callout-note
#### Code Explanation

Technically, it uses `select()`, `rename()`, `left_join()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. It defines `prepare_overrides()`, which encapsulates repeated logic used by later chunks. The iteration step loops across years, states, datasets, categories, or demographic groups and binds the repeated results into a single analysis table. The main objects created or updated here are `prepare_overrides`, `care_override`, `data`, `age_range`, `care_focus`.
:::


### Calculate Needs - Using ATUS

The next section in this document estimates the care needs of individuals (excluding child/elder care) for individuals who live alone. To undserstand how much care a person needs we look at individuals who live alone and provided no child or elder care for the day in their responses. We assume that these individuals are only providing care to themselves and thus the amount of time spent on caregiving represents the amount of care they demand. This assumption is likely somewhat flawed as it is possible that these people are providing care for others who are not in the household, such as baking food for a family member to be delivered later, but for the most part these activities are likely to be conducted for the indivdual doing them and no one else. We onlyy include activities that were conducted alone to try to limit activities that might be done to benefit a non-household member.

```{r market datum data_wrangling_05, message=FALSE, warning=FALSE}
age <- age_modified$age
needs_atus_calc <- list()

for (a in age) {
  # 1) Restrict to individuals with NO caregiving (primary or secondary) anywhere in their diary
  data <- atus |>
    group_by(caseid) |>
    filter(all(scc_all_ln == 0)) |>
    filter(all(sec_all_ln == 0)) |>
    filter(all(child_care == 0)) |>
    filter(all(elder_care == 0)) |>
    ungroup() |>
    # Keep care activities only (exclude "non-care")
    filter(care_focus != "non-care") |>
    # Living alone (household size = 1) and reference person
    filter(hh_size == 1) |>
    filter(relatew == 100) |>
    # Stabilize with a ±2-year window around age a
    filter(age %in% c(a - 2, a - 1, a, a + 1, a + 2))

  # 2) Aggregate per person, then compute weighted mean by care_focus
  data <- data |>
    reframe(
      duration = sum(duration, na.rm = TRUE),  # total minutes per person
      weight   = first(weight),                # person-level daily weight
      .by = c(care_focus, caseid)
    ) |>
    reframe(
      need_interval = weighted.mean(duration, w = weight, na.rm = TRUE),
      .by = c(care_focus)
    ) |>
    mutate(age = a)

  if (nrow(data) > 0) needs_atus_calc[[a]] <- data
}

needs_atus_calc <- bind_rows(needs_atus_calc)
```

::: callout-note
#### Code Explanation

Technically, it uses `filter()`, `mutate()`, `group_by()`, `bind_rows()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The iteration step loops across years, states, datasets, categories, or demographic groups and binds the repeated results into a single analysis table. The statistical work in this chunk calculates weighted counts, rates, rolling values, medians, changes over time, or inequality measures used in the published indicators. The main objects created or updated here are `age`, `needs_atus_calc`, `data`, `duration`, `weight`.
:::


### KU-Defined Overrides - Needs

As we previously mentioned, we will override certain peaces of data with predefined assumptions. The cod chunk below creates these assumptions. We use assumptions for children under the age of 18 because the ATUS only surveys individuals aged 15 and older. We also lack many data responses that fit our criteria of people who live alone for people between the age of 15 and 17 meaning we need to create estimates for this group as well.

For children under the age of 12, We assume that they need 24 hours of caregiving. This assumption comes from many state laws that require that a child is under care for 24 hours a day and it is a crime to leave them alone for any longer than that. For individuals over the age of 12 but still younger than 18 we think they need some moderate amount of care but less than 24 hours a day. We therefore assume a lower amount of careneeds for this group but one that is still higher than for 18-20 year olds.

We also provide overides for those over the age of 75. While we do have data for individuals over 75 in our data, we believe that the average person over the age of 75 who lives alone and receives no care from anyone else is likely different from the average elderly person. We therefore overide this data increasing the amount of care needed.

At the end of the day, these are assumptions. We believe they are valid assumptions but it is important to remember that these assumptions can be subject to change and people replicating this code are welcome to inser their own assumptions.

```{r market datum chunk_06, message=FALSE, warning=FALSE}
health_0to5 <- 300
develop_0to5 <- 420
daily_0to5 <- 1440 - health_0to5 - develop_0to5

health_6to12 <- 150
develop_6to12 <- 480
daily_6to12 <- 1440 - health_6to12 - develop_6to12

health_13to18 <- 90
develop_13to18 <- 360
daily_13to18 <- 1200 - health_13to18 - develop_13to18

health_75to84 <- 200
develop_75to84 <- 0
daily_75to84 <- 780 - health_75to84 - develop_75to84

health_85 <- 300
develop_85 <- 0
daily_85 <- 1200 - health_85 - develop_85

age_ranges <- bind_rows(
  expand.grid(age_range = "age_0to5", age = c(seq(0, 5, by = 1))),
  expand.grid(age_range = "age_6to12", age = c(seq(6, 12, by = 1))),
  expand.grid(age_range = "age_13to18", age = c(seq(13, 18, by = 1))),
  
  expand.grid(age_range = "age_75to84", age = c(seq(75, 84, by = 1))),
  expand.grid(age_range = "age_85", age = 85)
) 
  
need_interval <- list(
  age_0to5 = c("developmental" = develop_0to5, 
               "health" = health_0to5, "daily_living" = daily_0to5),
  
  age_6to12 = c("developmental" = develop_6to12, 
                "health" = health_6to12, "daily_living" = daily_6to12),
  
  age_13to18 = c("developmental" = develop_13to18, 
                 "health" = health_13to18, "daily_living" = daily_13to18), 
  
  age_75to84 = c("developmental" = develop_75to84, 
                 "health" = health_75to84, "daily_living" = daily_75to84),
  
  age_85 = c("developmental" = develop_85, 
             "health" = health_85, "daily_living" = daily_85)
)

needs_ku_override <- prepare_overrides(need_interval, age_ranges, "need_override")
```

::: callout-note
#### Code Explanation

Technically, it uses `bind_rows()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `health_0to5`, `develop_0to5`, `daily_0to5`, `health_6to12`, `develop_6to12`.
:::


### Calculate Provision - Using ATUS

Just as the ATUS data was utilized to calculate the need of a specific person, we can utilize it to calculate the average amount of care provided by each age group. Calculating care provision requires fewer assumptions and methodological wranging. Instead of needing to filter the data to only include care done by oneself, we want to know any care provided by an individual. Thus we simply utilize the columns occu_care_focus to look at paid care work and care_focus to look at unpaid care work and look at the time in each. This gives us an estimate about the amount of care provided on average by a given age group.

```{r market datum data_wrangling_07, message=FALSE, warning=FALSE}
age <- age_modified$age 
provision_atus_calc <- list()

for(a in age){
  data <- bind_rows(
    # formal care work
    atus |>
      filter(activity == "Formal Work" & occ_care_focus != "none") |> 
# Transform/add columns
      mutate(
        care_focus = occ_care_focus, 
        duration = duration 
        ),
    
    # informal care 
    atus |> filter(care_focus != "non-care"),

    # secondary child care - remove double counting
    atus |> 
      filter(scc_all_ln > 0) |>
      filter(!(activity == "Formal Work" & occ_care_focus != "none")) |> 
      filter(!(care_focus != "non-care")) |> 
      filter(sec_all_ln == 0) |>
# Transform/add columns
      mutate(
        care_focus = "developmental",
        duration = scc_all_ln
        ),
    
    # secondary elder care - remove double counting
    atus |> 
      filter(sec_all_ln > 0) |>
      filter(!(activity == "Formal Work" & occ_care_focus != "none")) |> 
      filter(!(care_focus != "non-care")) |> 
      filter(scc_all_ln == 0) |>      
# Transform/add columns
      mutate(
        care_focus = "health", 
        duration = sec_all_ln
        )
    )
  
  data <- data|> 
    filter(age %in% c(a-2, a-1, a, a+1, a+2)) %>%
    reframe(
      duration = sum(duration, na.rm = TRUE),
      weight = first(weight),
      .by = c(caseid, care_focus)
      ) %>%
    reframe( 
      provision_interval = wtd.quantile(
        duration, weights = weight, probs = 0.5, normwt = FALSE), 
      .by = care_focus
      ) |> 
# Transform/add columns
    mutate(age = a)

  if (nrow(data) > 0) {
    provision_atus_calc[[a]] <- data
  }
}
```

::: callout-note
#### Code Explanation

This chunk iterates over each age in the market table, builds a five-year age window around that age, combines formal work, primary informal care, secondary child care, and secondary eldercare into one activity table, and calculates weighted median provision minutes by care focus.
:::

```{r, message=FALSE, warning=FALSE}
provision_atus_calc <- bind_rows(provision_atus_calc)
```

::: callout-note
#### Code Explanation

This chunk collapses the list of age-specific provision estimates into a single `provision_atus_calc` data frame that can be joined to the market table and adjusted by the override rules in later chunks.
:::


### KU Defined Overrides

Just as with care need assumptions, we need to make assumptions about the amount of care provided by different age groups. We take a rather strict assumption here and start by saying that no one under the age of 18 provides any care and that no one over the age of 85 provides any care. This is clearly a flawed assumption, and in a coming subsection we will discuss how we relax this assumption a little to smooth out the estimates. But for now, we create these assumptions and then use the overide function to overide the blank estimates.

```{r market datum chunk_08, message=FALSE, warning=FALSE}
health_0to18 <- 0
develop_0to18 <- 0
daily_0to18 <- 0

health_85 <- 0
develop_85 <- 0
daily_85 <- 0

provision_interval <- list(
  age_0to18 = c("developmental" = develop_0to18, 
               "health" = health_0to18, "daily_living" = daily_0to18),
  
  age_85 = c("developmental" = develop_85, 
             "health" = health_85, "daily_living" = daily_85)
)

age_ranges <- bind_rows(
  expand.grid(age_range = "age_0to18", age = c(seq(0, 18, by = 1))),
  expand.grid(age_range = "age_85", age = 85)
) 

provision_ku_override <- prepare_overrides(provision_interval, age_ranges, "provision_override")
```

::: callout-note
#### Code Explanation

Technically, it uses `bind_rows()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `health_0to18`, `develop_0to18`, `daily_0to18`, `health_85`, `develop_85`.
:::


### Combine Need and Provision

Once we have our estimates of care need and care provision, we combine them together into a single dataframe.

```{r market datum data_wrangling_09, message=FALSE, warning=FALSE}
df_list <- list(
  market_datum,
  needs_atus_calc, 
  needs_ku_override, 
  provision_atus_calc, 
  provision_ku_override
) 

market_datum2 <- reduce(df_list, full_join, by = c("age", "care_focus")) |> 
# Transform/add columns
  mutate(
    need_interval = coalesce(need_override, need_interval, 0),
    provision_interval = coalesce(provision_override, provision_interval, 0)
  ) |> 
  select(age, care_focus, need_interval, provision_interval)
```

::: callout-note
#### Code Explanation

Technically, it uses `mutate()`, `select()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `df_list`, `market_datum2`, `need_interval`, `provision_interval`.
:::


### Plot Initial Market

We visualize the results of this analysis. As can be seen these results are very disjointed. This is due to the relativly small sample size as well as the assumptions we careate causing significant disjoints in the flow of the data.

```{r market datum visualization_10, message=FALSE, warning=FALSE}
# Generate diagnostic/summary plot
plot_market <- function(df){
  df_total <- df |> 
# Group data for aggregation
    group_by(age) |> 
# Summarise grouped data
    summarise(
      total_need = sum(need_interval)/60, 
      total_provision = -sum(provision_interval)/60
      )
  
    need <- comma(round(sum(df_total$total_need)))
    provision <- comma(round(sum(df_total$total_provision)))
    sel_title <- paste0("total need: ", need, "  total provision: ", provision)
  

  p <- ggplot(data = df_total, aes(x = age)) +
    geom_col(aes(y = total_need), fill = "grey") +
    geom_col(aes(y = total_provision), fill = "steelblue") +
    theme_minimal() +
    theme(axis.title = element_blank()) +
    ggtitle(sel_title)
  
  print(p)
}

# Generate diagnostic/summary plot
plot_market(market_datum2)
```

::: callout-note
#### Code Explanation

Technically, it uses `group_by()`, `summarise()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. It defines `plot_market()`, which encapsulates repeated logic used by later chunks. The statistical work in this chunk calculates weighted counts, rates, rolling values, medians, changes over time, or inequality measures used in the published indicators. The display code builds a diagnostic plot, publication figure, static table, or interactive HTML table from the objects created earlier.
:::


### Smoothing Function

In order to address the significant disjoints, we utilizing a smoothing function that will smooth the data by utilizing the surrounding values of an age group to learn and estimate adjustments. Ultimatly the reason for this smoothing function is to make the plots look more succint, however, it also has the value of allowing us to smooth over our assumptions. Now, instead of having hard disjoints in the data where we utilize our assumptions to overide the flow, we can use this smoothing algorithim to create a smooth transition across age groups. This leads to some of our overides being overidden with more appropriate data which allows our assumptions to be softer.

```{r market datum visualization_11, message=FALSE, warning=FALSE}
#Define function for smoothing while keeping within bounds
smooth_data <- function(df) {
  df %>%
# Group data for aggregation
    group_by(care_focus) %>%
    arrange(age) %>%  
# Transform/add columns
    mutate(
      smoothed_need = predict(
        loess(need_interval ~ age, data = cur_data(), span = 0.2), 
        newdata = data.frame(age = age)), 
      min_val_need = min(need_interval, na.rm = TRUE),
      max_val_need = max(need_interval, na.rm = TRUE),
      smoothed_need = pmax(pmin(smoothed_need, max_val_need), min_val_need), 

      smoothed_prov = predict(
        loess(provision_interval ~ age, data = cur_data(), span = 0.3), 
        newdata = data.frame(age = age)), 
      min_val_prov = min(provision_interval, na.rm = TRUE),
      max_val_prov = max(provision_interval, na.rm = TRUE),
      smoothed_prov = pmax(pmin(smoothed_prov, max_val_prov), min_val_prov) 
    ) %>%
    ungroup()
}

# Apply smoothing function to your dataset
market_datum_smoothed <- smooth_data(market_datum2) %>%
  select(age, care_focus, smoothed_need, smoothed_prov) %>%
  rename("need_interval" = smoothed_need) %>%
  rename("provision_interval" = smoothed_prov)

# Generate diagnostic/summary plot
plot_market(market_datum_smoothed)
```

::: callout-note
#### Code Explanation

Technically, it uses `mutate()`, `select()`, `rename()`, `group_by()`, `arrange()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. It defines `smooth_data()`, which encapsulates repeated logic used by later chunks. The main objects created or updated here are `smooth_data`, `smoothed_need`, `newdata`, `min_val_need`, `max_val_need`.
:::


### Final Overrides

While the smoothign algorithin does go over our assumptions, we want to reinstigate what we call our hard assumptions. These relate to need and provision for certain groups that we believe our assumptions were more valid for than the smoothed data. For example, the smoothed function shows that people under the age of 5 provide around 10 minutes of care in a day. We believe this is not accurate and thus re-overide the data for groups like this utilizing our assumptions and overide function.

```{r market datum chunk_12, message=FALSE, warning=FALSE}
health_0to5 <- 300
develop_0to5 <- 420
daily_0to5 <- 1440 - health_0to5 - develop_0to5

health_6to12 <- 150
develop_6to12 <- 480
daily_6to12 <- 1440 - health_6to12 - develop_6to12

#health_85 <- 300
#develop_85 <- 0
#daily_85 = 1200 - health_85 - develop_85

need_interval <- list(
  age_0to5 = c("developmental" = develop_0to5, 
               "health" = health_0to5, "daily_living" = daily_0to5),
  
  age_6to12 = c("developmental" = develop_6to12, 
                "health" = health_6to12, "daily_living" = daily_6to12)
)
  
#  age_85 = c("developmental" = develop_85, 
#             "health" = health_85, "daily_living" = daily_85)
#)

age_ranges <- bind_rows(
  expand.grid(age_range = "age_0to5", age = c(seq(0, 5, by = 1))),
  expand.grid(age_range = "age_6to12", age = c(seq(6, 12, by = 1))))
              
              
              
expand.grid(age_range = "age_85", age = 85)

final_need_override <- prepare_overrides(need_interval, age_ranges, "need_override")
```

::: callout-note
#### Code Explanation

Technically, it uses `bind_rows()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `health_0to5`, `develop_0to5`, `daily_0to5`, `health_6to12`, `develop_6to12`.
:::


```{r market datum chunk_13, message=FALSE, warning=FALSE}
no_provision <- 0

age_ranges <- bind_rows(
  expand.grid(age_range = "age_0to5", age = c(seq(0, 5, by = 1)))
) 

provision_interval <- list(
  age_0to5 = c("developmental" = no_provision, 
               "health" = no_provision, "daily_living" = no_provision)
)

final_provision_override <- prepare_overrides(provision_interval, age_ranges, "provision_override")
```

::: callout-note
#### Code Explanation

Technically, it uses `bind_rows()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. The main objects created or updated here are `no_provision`, `age_ranges`, `provision_interval`, `age_0to5`, `final_provision_override`.
:::


### Care Market

Finally, we join the different dataframes that we need to analyze together into a single market_datum file that we export as a csv. This csv will then be utilized to load data direction into the care board app.

```{r market datum visualization_14, message=FALSE, warning=FALSE}
df_list <- list(
  market_datum_smoothed,
  final_need_override,
  final_provision_override
) 

final_market_datum <- reduce(df_list, full_join, by = c("age", "care_focus")) |> 
# Transform/add columns
  mutate(
    need_interval = coalesce(need_override, need_interval, 0),
    provision_interval = coalesce(provision_override, provision_interval, 0)
  ) |> 
  select(age, care_focus, need_interval, provision_interval) |> 
  arrange(age, care_focus)

# Generate diagnostic/summary plot
plot_market(final_market_datum)

# Write output for the app and replication archive
write_careboard_outputs(
  final_market_datum,
  "market_datum",
  app_path = "./app_data/market_datum.csv"
)
```

::: callout-note
#### Code Explanation

Technically, it uses `mutate()`, `select()`, `arrange()` to restrict the analytic universe, create derived variables, reshape records, join lookup tables, and/or aggregate weighted statistics. It writes the resulting table through the shared export helper, producing app-facing files and/or replication copies in CSV, Stata, and Excel formats. The main objects created or updated here are `df_list`, `final_market_datum`, `need_interval`, `provision_interval`, `app_path`.
:::
